---
title: "bruder_module05_project01"
author: "Geoffrey Bruder"
date: "2024-10-14"
output:
  pdf_document: 
    latex_engine: xelatex
    toc: true
    toc_depth: 3
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
editor_options:
  markdown:
    wrap: 72
---

\newpage

------------------------------------------------------------------------

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#  Global Data Science Salaries: Navigating Wide Ranges and Remote Work Considerations to Attract Top Talent in a Competitive Market

A visual analysis of 
Salary Distribution: U.S.A. versus Offshore
Average Salary by Experience Level: U.S.A. versus Offshore
Salary versus Remote Work Ratio

Prepared by:  Geoffrey Bruder


```{r}
# Load necessary libraries
library(tidyr)
library(ggplot2)
library(dplyr)
library(readr)
library(knitr)
```


```{r}
# Read the data
data_path <- "/Users/Main/Documents/DSE5002/r project data.csv"
data <- read_csv(data_path)


# Filter data for full-time positions
full_time_data <- data %>% filter(employment_type == "FT")

# Summarize the full_time_data dataframe
summary(full_time_data)

# Display the first few rows of the full_time_data dataframe
head(full_time_data)

# List the column names of the full_time_data dataframe
colnames(full_time_data)

# Convert the column names of full_time_data dataframe to a list
colnames_list <- as.list(colnames(full_time_data))
```

```{r}
# Summary statistics for U.S. vs. offshore salaries
us_salaries <- full_time_data %>% filter(employee_residence == "US")
offshore_salaries <- full_time_data %>% filter(employee_residence != "US")

# Summary stats
us_summary <- us_salaries %>% summarize(
  avg_salary = mean(salary_in_usd, na.rm = TRUE),
  median_salary = median(salary_in_usd, na.rm = TRUE),
  sd_salary = sd(salary_in_usd, na.rm = TRUE)
)

offshore_summary <- offshore_salaries %>% summarize(
  avg_salary = mean(salary_in_usd, na.rm = TRUE),
  median_salary = median(salary_in_usd, na.rm = TRUE),
  sd_salary = sd(salary_in_usd, na.rm = TRUE)
)
```

```{r}
# Reformat the avg_salary, median_salary, and sd_salary columns
us_summary <- us_summary %>%
  mutate(
    avg_salary = paste0("$", formatC(avg_salary, format = "f", digits = 2)),
    median_salary = paste0("$", formatC(median_salary, format = "f", digits = 2)),
    sd_salary = paste0("$", formatC(sd_salary, format = "f", digits = 2))
  )

# View the reformatted dataframe
print(us_summary)

```

```{r}
# Reformat the avg_salary, median_salary, and sd_salary columns
offshore_summary <- offshore_summary %>%
  mutate(
    avg_salary = paste0("$", formatC(avg_salary, format = "f", digits = 2)),
    median_salary = paste0("$", formatC(median_salary, format = "f", digits = 2)),
    sd_salary = paste0("$", formatC(sd_salary, format = "f", digits = 2))
  )

# View the reformatted dataframe
print(offshore_summary)


```

#  Key Salary Metrics for Analyzing Compensation Data

```{r}
cat("Based on the analysis:  
    For U.S.-based data scientists: 
    average salary:", us_summary$avg_salary, 
    "median salary:", us_summary$median_salary,  
    "standard deviation", us_summary$sd_salary, 
     
    "For Offshore-based data scientists:, 
    average salary:", offshore_summary$avg_salary, 
    "median salary:", offshore_summary$median_salary,
    "standard deviation:", offshore_summary$sd_salary
)
```

#  Boxplot for Salary Distribution 

There is a wide variation in salary distributions across different residences.
Some residences have higher median salaries compared to others.
The U.S. shows a relatively high median salary and a wide range of salaries, including several outliers.
Other countries also show significant variations, with some having more outliers indicating higher salaries.
Disparities in salary distributions could be influenced by factors such as cost of living, demand for skills, and economic conditions in different regions.
Nonetheless, as tool for analyzing and understanding global salary trends, this visualization provides a clear comparison of salary distributions across various employee residences.

X-axis: Employee Residence (various country codes)
Y-axis: Salary in USD (ranging from 0 to 600,000 USD)
Boxes: Represent the interquartile range (IQR), which contains the middle 50% of the data.
Horizontal Line Inside the Box: Represents the median salary for that residence.
Whiskers: Extend from the boxes to the smallest and largest values within 1.5 times the IQR from the quartiles.
Dots: Represent outliers, which are data points outside the whiskers.

```{r}
# Boxplot for Salary Distribution with adjusted margins
ggplot(full_time_data, aes(x = employee_residence, y = salary_in_usd)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Salary Distribution: U.S. vs Offshore",
       x = "Employee Residence",
       y = "Salary in USD") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8)) +
  theme(plot.margin = unit(c(1,1,1,2), "cm")) %>% 
  print()
```

#  Bar graph for Average Salary by Experience Level (U.S. versus Offshore)

The bar graph provides a clear visual comparison of average salaries for different experience levels across U.S. and Offshore locations.
U.S.-based employees have higher average salaries compared to their offshore counterparts across all experience levels.
This highlights potential salary gaps and can inform strategic decisions regarding competitive compensation offers by emphasizing the importance of considering geographic location and experience level in salary planning.
X-Axis: Experience Level (with categories: Entry Level, Mid-Level, Senior Level, Executive Level)
Y-Axis: Average Salary in USD

```{r}
# Bar plot for Average Salary by Experience Level in U.S. and Offshore
avg_salary_experience <- full_time_data %>%
  group_by(employee_residence, experience_level) %>%
  summarize(avg_salary = mean(salary_in_usd, na.rm = TRUE)) %>%
  ungroup()
```

```{r}

ggplot(avg_salary_experience, aes(x = experience_level, y = avg_salary, fill = employee_residence)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(title = "Average Salary by Experience Level: U.S. vs Offshore",
       x = "Experience Level",
       y = "Average Salary in USD",
       fill = "Employee Residence") +
  theme(plot.title = element_text(size = 10, hjust = 0.5)) %>% 
  print()

```

# Scatter plot for Salary versus Remote Work Ratio

The scatter plot represents the relationship between salary in USD and remote work ratio for employees from different countries, emphasizing the salary distribution. 
Salaries vary widely across all remote ratios. 
Some employees in the fully remote category (100% remote ratio) are earning higher salaries, indicating that remote work may be associated with more competitive compensation for certain roles in in certain regions.

X-Axis (Remote Ratio): This axis represents the percentage of work that employees perform remotely, ranging from 0% (fully on-site) to 100% (fully remote).
Y-Axis (Salary in USD): This axis shows the salaries of the employees, measured in US dollars.
Color (Employee Residence): Each color represents a different country where the employees reside.

```{r}
# Scatter plot for Salary versus Remote Work Ratio
ggplot(full_time_data, aes(x = remote_ratio, y = salary_in_usd, color = employee_residence)) +
  geom_point() +
  theme_minimal() +
  labs(title = "Salary vs Remote Ratio",
       x = "Remote Ratio",
       y = "Salary in USD",
       color = "Employee Residence") %>%
  print()
```

#  Strategic Salary Recommendations

Balancing U.S. and Offshore Data Science Compensation in a Competitive Market


```{r}
# Conclusion based on the analysis
cat("Based on the analysis, I recommend the following:  
    For U.S.-based data scientists, 
    the average salary is around", us_summary$avg_salary, 
    "USD, 
    while offshore data scientists have an average 
    salary of approximately", offshore_summary$avg_salary,"USD.
    To attract top talent, especially in a competitive market, 
    consider offering salaries at or above these averages 
    depending on the candidate's experience and expertise.
    Considering a notable relationship between the remote 
    work ratio and salary distribution for employees across 
    different countries, as well as, current trends in the 
    influence of remote work on hiring salary, more 
    competitive compensation may be necessary for roles 
    allowing full remote work.  
    
    These considerations may help set competitive salaries 
    based on average salaries, remote work ratios, 
    and geographical locations. It can also inform decisions 
    about remote work policies and assist in developing 
    guidelines and policies that support fair compensation 
    practices.")
```





```{r}
company_data <- data %>% filter(employment_type == "FT")
# Summarize the full_time_data dataframe
summary(full_time_data)

# Display the first few rows of the full_time_data dataframe
head(full_time_data)

# List the column names of the full_time_data dataframe
colnames(full_time_data)
```

```{r}
# Filter data for full-time employees and calculate the averages
average_remote_work <- company_data %>%
  filter(employment_type == "FT") %>%
  group_by(company_location, job_title, remote_ratio) %>%
  summarise(average_percentage = mean(as.numeric(remote_ratio), na.rm = TRUE) * 100) %>%
  ungroup()


```


```{r}
# Filter data for full-time employees and calculate the averages
average_remote_work <- company_data %>%
  filter(employment_type == "FT") %>%
  mutate(group = ifelse(company_location == "US", "US", "Offshore")) %>%
  group_by(group, company_size, remote_ratio) %>%
  summarise(average_percentage = n() / nrow(company_data) * 100) %>%
  ungroup()

# Reshape the data for better readability
average_remote_work_table <- average_remote_work %>%
  pivot_wider(names_from = remote_ratio, values_from = average_percentage, names_prefix = "Remote_") %>%
  rename(Remote_0 = `Remote_0`, Remote_50 = `Remote_50`, Remote_100 = `Remote_100`) %>%
  arrange(group, company_size)

```

#  Average Percentage of Full-Time Employees Working Remotely by Company Size and Group (US versus Offshore)

The data represented in chart indicates that US companies, especially medium-sized ones, have a higher percentage of full-time employees working remotely compared to Offshore companies.

Remote work trends across different company sizes and geographical locations highlight the variations in the adoption of remote work between large, medium, and small companies in the US and Offshore. 

```{r}
# Display the data as a table
kable(average_remote_work_table, caption = "Average Percentage of Full-Time Employees Working 
      Remotely by Company Size and Group (US/Offshore)",
      col.names = c("Group", "Company Size", "0% Remote", "50% Remote", "100% Remote"))

```

```{r}
#{r echo=FALSE, message=FALSE, warning=FALSE, results='hide'}

```

